A load balancing scheme for MMOGs

### 2009-03-31 textos/dissertation/MTAP_LaTeX/pub/tex/latex/svjour3/global/balancing.tex

An initial idea might be to distribute the players between servers, so that the number of players on each server would be proportional to the bandwidth of that server. However, this distribution would not work, for the burden caused by players also depends on how they are interacting with one another. For example, if the avatars of two players are too distant from each other, probably there will be no interaction between them and therefore the server needs only to update each of the result of their own actions. However, if these avatars are close to each other, each player should be updated not only of his own actions, but also of the actions of the other player.

Normally, players can freely move their avatars throughout the game world. This makes possible the formation of \emph{hotspots}, around which the players are more concentrated than in other regions of the virtual environment. Moreover, many massively multiplayer online RPGs not only permit but also stimulate, to some extent, the formation of these points of interest. In the worlds of these MMORPGs, there are entire cities, where the players meet to chat, exchange virtual goods or even fight, and there are also desertic areas, with few attractions for the players, and where the number of avatars is relatively small compared to other places of the environment.

@@

%**	
To tackle the problem, the authors also subdivide the virtual environment in rectangular \textbf{cells} -- similar to the microcrells. The cells are grouped in \textbf{regions} -- or macrocells -- and each region is managed by one server. Each server keeps its players updated of the state of the virtual environment, while also handling the interactions between avatars in its region. Two cells are adjacent (or neighbors) if they share a border. Similarly, two regions -- and their servers -- are adjacent if there is a pair of adjacent cells, each of which belong to one of the two regions. Figure \ref{fig:leeA} illustrates a distributed server system, consisting of 16 nodes. The virtual environment is divided in 256 cells, which are grouped in 16 regions.

%**
\begin{figure}
 \centering
 \includegraphics[width=0.8\textwidth]{images/LEE_leeA.jpg}
 \caption{Cells and regions in a distributed server system}
 \label{fig:leeA}
\end{figure}

@@

One more important point that was not properly considered by previous works, is that the main criterion to be considered when balancing the of MMOG servers is the bandwidth, and not much processing power. Several games include simulations of virtual environments of hundreds or thousands of entities, such as the game Age of Empires, which are performed without problems in personal computers today. However, if this game were multiplayer and each of these entities were controlled by a player in a network, connected to other players, it would most likely generate a traffic ammount which would hardly be supported by a domestic connection.

Moreover, the upload bandwidth must be taken into account, much more than download. This happens for two reasons: first, the usage of the download bandwidht of each server node grows linearly with the number of players connected to it, while the usage of the upload bandwidth may have a quadratic growth, getting quickly overwhelmed; second, the domestic connections -- which are majority in a volunteer peer-to-peer system -- usually have an upload bandwidth much smaller than the download one.

@@

The choice of cells to include in the region tries to put the most interacting pairs of cells in the same regions, reducing the overhead.% In each step, the algorithm not only tries to get a neighbor cell, but also the one with the highest that not only is free from a nearby cell already present in the region, but whose edge linking them is the heaviest possible. %Figure [TODO: reffig] shows an example of the steps of the growth of a region to achieve its share of the load balancing.

In the first step of the cycle starting in line \ref{alg:ggp:whilefraction}, if the region does not yet have any cell, \ggp\ gets the heaviest free cell (line \ref{alg:ggp:freeheavycell}). The same occurs when a region is compressed between the borders of other regions and has no free neighbor, getting cells from somewhere else. This may generate fragmented regions and possibly increase the overhead of the game. However, this happens more often in the last steps of the distribution, when most of the cells would be already allocated to some region. Because the algorithm is greedy, when it reaches that stage of its execution, the free cells would probably the lightest cells of the environment, causing little overhead.

>>>>

An initial idea might be to distribute the players between servers, so that the number of players on each server would be proportional to the bandwidth of that server. However, this distribution would not work, for the burden caused by players also depends on how they are interacting with one another. For example, if the avatars of two players are too distant from each other, probably there will be no interaction between them and therefore the server needs only to update each one of them with the result of their own actions. However, if these avatars are close to each other, each player should be updated not only of his own actions, but also of the actions of the other player.

Normally, players can freely move their avatars throughout the game world. This makes possible the formation of \emph{hotspots} \cite{ahmed2008mol}, around which the players are more concentrated than in other regions of the virtual environment (Figure \ref{fig:avatarsdistribution}). Moreover, many massively multiplayer online RPGs not only permit but also stimulate, to some extent, the formation of these points of interest. In the worlds of these MMORPGs, there are entire cities, where the players meet to chat, exchange virtual goods or even fight, and there are also desertic areas, with few attractions for the players, and where the number of avatars is relatively small compared to other places of the environment.

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{images/avatarsdistribution}
  \caption{Distribution of avatars with and without hotspots}
  \label{fig:avatarsdistribution}
\end{figure}

@@

%**	
%To tackle the problem, the authors also subdivide the virtual environment in rectangular \textbf{cells} -- similar to the microcrells. The cells are grouped in \textbf{regions} -- or macrocells -- and each region is managed by one server. Each server keeps its players updated of the state of the virtual environment, while also handling the interactions between avatars in its region. Two cells are adjacent (or neighbors) if they share a border. Similarly, two regions -- and their servers -- are adjacent if there is a pair of adjacent cells, each of which belong to one of the two regions. Figure \ref{fig:leeA} illustrates a distributed server system, consisting of 16 nodes. The virtual environment is divided in 256 cells, which are grouped in 16 regions.

%Para atacar o problema, os autores também subdividem o ambiente virtual em \textbf{células} -- semelhantes às microcélulas -- retangulares, sendo que o número de servidores é muito menor que o número de células. As células são agrupadas em \textbf{regiões} -- ou macrocélulas -- e cada região é gerenciada por um servidor. Cada servidor mantém atualizadas as informações de estado dos usuários e lida com as interações entre avatares na região a ele dedicada. Cada usuário envia e recebe atualizações de estado através do servidor que gerencia a região na qual ele está jogando. Duas células são ditas adjacentes (ou vizinhas) se elas compartilharem uma fronteira. Analogamente, duas regiões, e seus respectivos servidores, são ditos adjacentes se existir um par de células adjacentes, cada uma das quais pertencendo a uma das duas regiões.

To tackle the problem, the authors also subdivided the virtual environment in rectangular \textbf{cells} -- similar to the microcells --, where the number of servers is much smaller than the number of cells. The cells are grouped into \textbf{regions} -- or macrocells -- and each region is managed by one server (Figure \ref{fig:macromicro}. Each server handles the interaction between avatars in the region dedicated to him. It receives the inputs of the players controlling these avatars and sends back to them the up-to-date game state. Two cells are called adjacent (or neighbors) if they share a border. Similarly, two regions, and the servers assigned to them, are called adjacent (neighbors) if there is a pair of adjacent cells, each of which belonging to one of the two regions.

\begin{figure}
 \centering
 \includegraphics[width=0.6\textwidth]{images/macromicro}
 \caption{Cell grouped into four regions (R$_\text{1}$, R$_\text{2}$, R$_\text{3}$ and R$_\text{4}$)}
 \label{fig:macromicro}
\end{figure}

%**
%\begin{figure}
% \centering
% \includegraphics[width=0.8\textwidth]{images/LEE_leeA.jpg}
% \caption{Cells and regions in a distributed server system}
% \label{fig:leeA}
%\end{figure}

@@

One more important point that was not properly considered by previous works, is that the main criterion to be considered when balancing the of MMOG servers is the bandwidth, and not much processing power. Several games include simulations of virtual environments of hundreds or thousands of entities, such as the game Age of Empires \cite{ageofempires}, which are performed without problems in personal computers today. However, if this game were multiplayer and each of these entities were controlled by a player in a network, connected to other players, it would most likely generate a traffic ammount which would hardly be supported by a domestic connection \cite{feng2007wnn}.

%TODO: explicar porque cresce linearmente
Moreover, the upload bandwidth must be taken into account, much more than download. This happens for two reasons: first, the usage of the download bandwidht of each server node grows linearly with the number of players connected to it, while the usage of the upload bandwidth may have a quadratic growth, getting quickly overwhelmed; second, the domestic connections -- which are majority in a volunteer peer-to-peer system -- usually have an upload bandwidth much smaller than the download one.

%TODO: translate
Há questões de menor importância, mas que também devem ser levadas em consideração e que não foram consideradas em outros trabalhos. Uma delas é que dificilmente o sistema servidor será homogêneo -- considerando-se que é baseado em recursos voluntários. Portanto, não se pode assumir que os servidores tenham a mesma quantidade de recursos. Outro problema, específico do algoritmo proposto por \cite{lee2003sdl}, é que o critério de alocar para cada servidor uma carga equivalente a 90\% da sua capacidade, com o fim de evitar rebalanceamentos constantes, é fraco. Não há garantias de que o sistema servidor terá 11,11\% a mais de capacidade total do que o necessário. Muito pelo contrário, ele deve se adaptar a situações de sobrecarga generalizada. Além disso, o critério de parada é fraco pois o algoritmo não termina se o sistema todo estiver com carga acima de 90\% da sua capacidade, ou se o servidor sobrecarregado não tiver vizinhos disponíveis para balanceamento.

There are also other issues which must not be overlooked. One of them is that the server system is probably not homogenous -- considering that it is based on volunteer resources. Therefore, one can not assume that the servers have the same amount of cpu power or network bandwidth. Another problem, specific to the algorithm proposed by \cite{lee2003sdl}, is that the criterion to allocate to each server a load equivalent to 90\% of its capacity in order to avoid constant rebalancing is weak. There are no guarantees that the server system has 11.11\% more total capacity than the necessary. On the contrary, it should adapt to situations of widespread overload. Moreover, the criterion for stopping is weak because the algorithm does not stop when the whole system is loaded above 90\% of its capacity, or if the overloaded server has no neighbors capable of receiving more load.

%TODO: translate
Porém, uma idéia importante que foi sugerida é a do uso de grafos para representar o ambiente virtual e de usar algoritmos de particionamento de grafo para realizar o balanceamento de carga. No capítulo \ref{cap:contexto}, foi feita uma introdução a respeito desse princípio, que é a base do esquema de balanceamento de carga aqui proposto. Na seção seguinte, será mostrado como um ambiente virtual de um MMOG pode ser mapeado em um grafo.

@@

%The choice of cells to include in the region tries to put the most interacting pairs of cells in the same regions, reducing the overhead.% In each step, the algorithm not only tries to get a neighbor cell, but also the one with the highest that not only is free from a nearby cell already present in the region, but whose edge linking them is the heaviest possible. %Figure [TODO: reffig] shows an example of the steps of the growth of a region to achieve its share of the load balancing.

%A escolha de células para incluir na região busca fazer com que cada uma das arestas mais pesadas do grafo $GW$ ligue vértices da mesma partição, reduzindo o corte de aresta e, assim, o overhead. A cada passo é dada preferência a escolher a célula livre que não apenas seja vizinha de uma célula já presente na região, mas também cuja aresta ligando-as seja a mais pesada possível. A Figura \ref{fig:ggp} mostra um exemplo dos passos do crescimento de uma região até atingir a sua parcela de carga no balanceamento.

The criterion to choose a cell to include in the region is to make the heaviest edges on the graph $GW$ connect vertices in the same partition, reducing the edge-cut and, thus, the overhead. In each step, it is selected the cell which not only is adjacent to a cell already present in the region, but whose edge connecting them is the heaviest possible. Figure \ref{fig:ggp} illustrates the steps of growth of a region until it reaches its load share.

%TODO: referenciar karypis e kumar, wow e outros jogos (eve-online, warhammer e cabal)
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{images/ggp}
\caption{Growth of a partition (region) with \ggp}
\label{fig:ggp}
\end{figure}

In the example of Figure \ref{fig:ggp}, there are two servers, $S_1$ and $S_2$, where $p(S_1) = 30$ and $p(S_2) = 18$. The total weight of the environment being repartitioned is $W_{total} = 32$. For the division to be proportional to the capacity of each server, the weights assigned to $S_1 $ and $S_2$ are 20 and 12, respectively. The selection starts with the vertex of weight 6 (free cell with the highest weight) and, after that, at each step the vertex connected by the heaviest edge is added to the partition. The selected edges and the vertices belonging to the new partition are highlighted.

In the first step of the cycle starting in line \ref{alg:ggp:whilefraction}, if the region does not have any cell yet, \ggp\ gets the heaviest free cell (line \ref{alg:ggp:freeheavycell}). The same occurs when a region is compressed between the borders of other regions and has no free neighbor, getting cells from somewhere else (Figure \ref{fig:compressed}). This may generate fragmented regions and possibly increase the overhead of the game. However, this happens more often in the last steps of the distribution, when most of the cells would be already allocated to some region. Because the algorithm is greedy, when it reaches that stage of its execution, the free cells would probably the lightest cells of the environment, causing little overhead.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{images/compressed}
\caption{Formation of disconnected partition (fragmented region)}
\label{fig:compressed}
\end{figure}
